---
title: "Codebook for UCI HAR dataset"
author: "Adrian"
date: "December 26, 2015"
output: html_document
---

Revision history:

```{r, echo=FALSE, warning=FALSE}
library(dplyr)
revisionhistory <- data.frame(matrix(ncol = 3, nrow = 1))
names(revisionhistory) <- c("Version", "Date", "Comments")
revisionhistory$Version[1] <- "1.0"
revisionhistory$Date[1] <- "2015/12/26"
revisionhistory$Comments[1] <- "First version of document"
revisionhistory
```
##Introduction
#####The following codebook describes the  variables, the data, and any transformations or work I performed to clean up the data in the Human Activity Recognition Using Smartphones Smartphones dataset, for the Getting and Cleaning Data class in Coursera.

#### First I read in the data. Variable names are pretty self-explanatory. I just removed the underscores.
```{r, echo=FALSE}
setwd("C:/Users/Adrian/Documents/R/coursera/getdata-035/UCI HAR Dataset/")
subjecttest <- read.table("subject_test.txt", header = FALSE)
subjecttrain <- read.table("subject_train.txt", header = FALSE)
features <- read.table("features.txt", header = FALSE)
activitylabels <- read.table("activity_labels.txt", header = FALSE)
xtest <- read.table("X_test.txt", header = FALSE)
xtrain <- read.table("X_train.txt", header = FALSE)
ytest <- read.table("y_test.txt", header = FALSE)
ytrain <- read.table("y_train.txt", header = FALSE)
```
### Task 1: 
####Merges the training and the test sets to create one data set.

I use rbind() to row-bind the xtrain and xtest sets, in that order. The new data is called xdata. I do the same (in the same order) for the Y data. Then I use cbind() to bind ydata and xdata. The new variable is called alldata. Task 1 is complete.
```{r, echo = FALSE}
xdata <- rbind(xtrain, xtest)
ydata <- rbind(ytrain, ytest)
alldata <- cbind(ydata, xdata)
```
### Task 2:
####Extracts only the measurements on the mean and standard deviation for each measurement. 
To do this, I want to rename the column names away from "V550" etc. to "features.txt". I want unique column names, but features has duplicate names, resulting in:
"Error: found duplicated column name: etc. etc."
So I use the make.names() function to make them into unique names in a new column in features. Thanks to Saeid Abolfazli in the forums for this.
However because I cbinded ydata and xdata into alldata, I need to put the name for the single columm of ydata in there first.
Then I can use the select() and contains() functions to get only the variables with names that contain "mean" or "std". That doesn't include the activity type from ydata, so I have to add that in too.
After this, alldata contains only the ydata, the data with means, and the data with standard deviations.
```{r, echo = FALSE}
features$V3 <- make.names(features$V2, unique = TRUE)
names(alldata) <- c("activitynum", features$V3)
alldata <- cbind(
  select(alldata, activitynum), 
  select(alldata, contains("mean")), 
  select(alldata, contains("std"))
  )
```
### Task 3:
#### Uses descriptive activity names to name the activities in the data set
To do this, I want to merge the activitylabels dataframe with my data. However merge re-orders the data, so I can only use it after I've rbinded/cbinded everything together.

```{r, echo = FALSE}
subject <- rbind(subjecttrain, subjecttest)
names(subject) <- "subject"
alldata <- cbind(subject, alldata)
names(activitylabels) <- c("num", "activity")
alldata <- merge(
  x = activitylabels, y = alldata, 
  by.x = "num", by.y = "activitynum", 
  all = TRUE)
```
### Task 4:
#### Appropriately labels the data set with descriptive variable names. 
I did this already in Task 2 with make.names().

### Task 5:
#### From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
To do this, I first delete the activitynum from the dataset (because it duplicates the activity name field), then group by activity and subject, then use the summarize_each() function. The new tidy data set is called "tidy."
```{r, echo = FALSE}
alldata <- subset(alldata, select = -num)
alldata <- group_by(alldata, activity, subject)
tidy <- summarize_each(alldata, funs(mean))
#ungroup(alldata)
```
### Final task:
#### Please upload your data set as a txt file created with write.table() using row.name=FALSE

```{r, echo = TRUE}
write.table(x = tidy, file = "tidydata.txt", row.names = FALSE)
```

### List of variables and their descriptions:

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

tBodyAcc-XYZ
tGravityAcc-XYZ
tBodyAccJerk-XYZ
tBodyGyro-XYZ
tBodyGyroJerk-XYZ
tBodyAccMag
tGravityAccMag
tBodyAccJerkMag
tBodyGyroMag
tBodyGyroJerkMag
fBodyAcc-XYZ
fBodyAccJerk-XYZ
fBodyGyro-XYZ
fBodyAccMag
fBodyAccJerkMag
fBodyGyroMag
fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

mean(): Mean value
std(): Standard deviation
meanFreq(): Weighted average of the frequency components to obtain a mean frequency
Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

gravityMean
tBodyAccMean
tBodyAccJerkMean
tBodyGyroMean
tBodyGyroJerkMean

```{r, echo = FALSE}

varNames <- data.frame(
  "Variable number" = seq(from = 1, to = ncol(alldata), by = 1),
  Names = names(alldata))
print(varNames)
```

